---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Desa Bolger"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
#install.packages("agricolae")
library(agricolae) 
library(tidyverse)
library(lubridate)
library(ggplot2)
library(here)
library(ggthemes)
library(viridis)
here()

ChemPhysics <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
                        stringsAsFactors = TRUE)

#view(ChemPhysics)


class(ChemPhysics$sampledate) #05/27/84
# a factor! we will switch it
ChemPhysics$sampledate <- as.Date(ChemPhysics$sampledate, format = "%m/%d/%y")
class(ChemPhysics$sampledate) #now its a date!
#view(ChemPhysics)

#2

desa_theme <- theme_base() + 
  theme(line = element_line(color='black',linewidth =.5),
        text = element_text(color='black'),
        panel.grid.major =  element_line(color='black', linewidth = .5),
        rect = element_rect(color = 'lightgrey', fill = 'lightgrey'),
        plot.background = element_rect(color = 'lightgrey', fill = 'lightgrey'),
        panel.background =  element_rect(color = 'lightgrey', fill = 'lightgrey'),
        legend.background = element_rect(color='lightblue', fill = 'lightblue'),
        legend.title = element_text(color='darkblue'))

theme_set(desa_theme) #set as default

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question: 
> Answer: 
H0: Mean lake temperature does not change with depth across all lakes.
HA: Mean lake temperature does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

ChemPhysics.processed <- 
  ChemPhysics %>%
  mutate("Month" = month(sampledate)) %>%
  filter(Month == 7) %>%  
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na()

#view(ChemPhysics.processed)


#5 

ChemGraph <- ggplot(ChemPhysics.processed, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = 0.5, size = 1.5) +
  stat_smooth(method = lm, se = F, color = 'red')+
  ylim(0,35)+
  xlab(expression(paste("Depth (m)" )))+
  ylab(expression(paste("Temperature (C)")))


print(ChemGraph) 


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: It suggests that as depth increases, temperature decreases.  The distribution seems to follow the line, suggesting it may be linear, but we should do a linear regression to be sure. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7 

Chem.regression <- 
  lm(ChemPhysics.processed$temperature_C ~ 
       ChemPhysics.processed$depth)

print(Chem.regression)
print("y = -19.46x + 21.956")

summary(Chem.regression)

cor.test(ChemPhysics.processed$temperature_C, ChemPhysics.processed$depth)
print("df = 9726. p value < 2.2e-16")

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The df is 9726, the p-value is <2.2e-16 which is less than 0.05.  This means it is statistically significant, so we can reject the null hypothesis.  
For every 1 m change in depth, temperature decreases by -19.46 degrees C.  
The R-squared is 0.7387, suggesting that much of the variablity of temperature 
is explained by this model.  The equation means that temperature can be described by taking the depth, multiplying by -19.46 and adding 21.956. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

CPaic <- lm(data = ChemPhysics.processed, temperature_C ~ year4 + daynum + depth)

#Choose a model by AIC in a Stepwise Algorithm
step(CPaic)


CPmodel <- lm(data = ChemPhysics.processed, temperature_C ~ year4 + daynum + depth)
summary(CPmodel)

#Best thing is to remove no variables, as this has the lowest AIC. The full model
#the best, so we should use all three variables.


#10

CPregression <- lm(data = ChemPhysics.processed, temperature_C ~ year4 + daynum + depth)
summary(CPregression)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: (CHECK)
year4, daynum, and depth are the three sets of explanatory variables that the AIC method suggests, as the probability for each is less than 0.05, meaning they are statistically significant. The R-squared is now 0.7412, which means the data is well-explained, but this is also the R-squared for the linear model, so that is not an improvement. 


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12 

# Format ANOVA as aov
Chem.anova <- aov(data = ChemPhysics.processed, temperature_C ~ lakename)
summary(Chem.anova)

#results: reject null hypothesis i.e. difference between a pair of group means 
#is statistically significant

# Format ANOVA as lm
Chem.anova2 <- lm(data = ChemPhysics.processed, temperature_C ~ lakename)
summary(Chem.anova)

#results: reject null hypothesis i.e. difference between a pair of group means 
#is statistically significant


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: yes. The P value is <2e-16 (less than 0.05), so it is statistically significant. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14. 

ggplot(ChemPhysics.processed, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5, ) +
  geom_smooth(method = "lm", se = FALSE, ) + 
  scale_color_viridis(discrete = TRUE, option = "turbo", name = "Lakes") +

  labs(x = "Depth (m)", y = "Temperature (C)")+
  ylim(0, 35)


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

TukeyHSD(Chem.anova)

ChemTukey <- HSD.test(Chem.anova, "lakename", group = TRUE)
ChemTukey

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: (CHECK) Ward, Peter, and Paul lakes have the same mean temperature statistically speaking because they are in the same group. All groups have at least two lakes, so there is not a lake that is statistically distinct from the others. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: (CHECK) 2 sample T-test



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
# ASK for answer

ChemPhysics.processed2 <- 
  ChemPhysics.processed %>%
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake") 
view(ChemPhysics.processed2)

CP.twosample <- t.test(ChemPhysics.processed2$temperature_C ~ 
                         ChemPhysics.processed2$lakename)
CP.twosample

```

>Answer: The p-value is 0.2649, which suggests the mean temperatures are not statistically significant in their difference.  The mean for Crampton is 15.35189 and the mean for Ward Lake is 14.45862. This suggests that they may actually have the same mean, which does match my answer in 16.  
