---
title: "Assignment 5: Data Visualization"
author: "Desa Bolger"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `Desa_Bolger_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 Getting things added!
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
library(ggplot2)
library(ggthemes)
library(viridis)
library(RColorBrewer)

here() #"/home/guest/EDE_Fall2023"

processed_data = "Data/Processed/Processed_KEY"
PeterPaul.chem.nutrients <- read.csv(
  here(processed_data,"NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
  stringsAsFactors = TRUE)

#view(PeterPaul.chem.nutrients) #Looks Good

Niwot_Ridge <- read.csv(
  here(processed_data,"NEON_NIWO_Litter_mass_trap_Processed.csv"), 
  stringsAsFactors = TRUE)
view(Niwot_Ridge)

#2 Change to Date!

class(PeterPaul.chem.nutrients$sampledate)
PeterPaul.chem.nutrients$sampledate <- ymd(PeterPaul.chem.nutrients$sampledate)
class(PeterPaul.chem.nutrients$sampledate)
class(Niwot_Ridge$collectDate)
Niwot_Ridge$collectDate <- ymd(Niwot_Ridge$collectDate)
class(Niwot_Ridge$collectDate)

#Date changed
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3 Building my plot theme

desa_theme <- theme_base() + 
  theme(line = element_line(color='black',linewidth =.5),
        text = element_text(color='black'),
        panel.grid.major =  element_line(color='black', linewidth = .5),
        rect = element_rect(color = 'lightgrey', fill = 'lightgrey'),
        plot.background = element_rect(color = 'lightgrey', fill = 'lightgrey'),
        panel.background =  element_rect(color = 'lightgrey', fill = 'lightgrey'),
        legend.background = element_rect(color='lightblue', fill = 'lightblue'),
        legend.title = element_text(color='darkblue'))

theme_set(desa_theme) #set as default

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 PHOS x PHOSPHATE.    

PhosByPhos <- ggplot(PeterPaul.chem.nutrients, aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point(alpha = 0.75, size = 1.5) +
  xlim(0,50)+
  scale_color_viridis(discrete = TRUE, option = "turbo", name = "Lakes") +
  stat_smooth(method = lm, se = F, color = 'black')+
  xlab(expression(paste("Phosphate" )))+
  ylab(expression(paste("Total Phosphorus (", mu, "g)")))

print(PhosByPhos) 

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5. Boxplot!!!    

PeterPaul.chem.nutrients$month <- as.factor(PeterPaul.chem.nutrients$month)

Temp <- ggplot(PeterPaul.chem.nutrients, aes(x = month, y = temperature_C, color = lakename)) +
  geom_boxplot(fill = 'lightgrey')+
  scale_x_discrete(limits = factor("5":"11"), labels = c("May", "June", "July", "Aug", "Sept", "Oct", "Nov"))+
  xlab(expression(paste("Month")))+
  ylab(expression(paste("Temperature (degrees C)")))+
  scale_color_brewer(palette = "Dark2", name = "Lakes") 

print(Temp) 

# getting the legend
FinalLegend <- get_legend(Temp)

TP <- ggplot(PeterPaul.chem.nutrients, aes(x = month, y = tp_ug, color = lakename)) +
  geom_boxplot(fill = 'lightgrey')+
  scale_x_discrete(limits = factor("5":"11"), labels = c("May", "June", "July", "Aug", "Sept", "Oct", "Nov"))+
  xlab(expression(paste("Month")))+
  ylab(expression(paste("Total Phosphorus (", mu, "g)")))+
  scale_color_brewer(palette = "Dark2", name = "Lakes") 

print(TP) 

TN <- ggplot(PeterPaul.chem.nutrients, aes(x = month, y = tn_ug, color = lakename)) +
  geom_boxplot(fill = 'lightgrey')+
  scale_x_discrete(limits = factor("5":"11"), labels = c("May", "June", "July", "Aug", "Sept", "Oct", "Nov"))+
  xlab(expression(paste("Month")))+
  ylab(expression(paste("Total Nitrogen (", mu, "g)")))+
  scale_color_brewer(palette = "Dark2", name = "Lakes") 

print(TN) 

#issue: font, not aligned, colored background?

finished <- plot_grid(
  Temp + theme(legend.position = "none"),
  TP + theme(legend.position = "none"),
  TN + theme(legend.position = "none"),
  FinalLegend,
  nrow = 2,
  ncol = 2,
  theme(axis.text = element_text(size = 6)), #not working
  align = 'hv')


print(finished)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Temperature is highest in the summer months (July and August), but fairly consistent for Paul and Peter Lakes. Total Phosphate is highest in July, and it seems Peter lake has a higher TP level than Paul.  Lastly, TN seems fairly consistent, though Peter lake values seem higher. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6 Isolating Needles.       

Needles <- ggplot(subset(Niwot_Ridge, functionalGroup == 'Needles' ), aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  scale_color_viridis(discrete = TRUE, name = "NLCD Class") +
  xlab(expression(paste("Year" )))+
  ylab(expression(paste("Dry Mass (g)"))) #check grams
  print(Needles) 

#7

Needles2 <- ggplot(subset(Niwot_Ridge, functionalGroup == 'Needles' ), aes(x = collectDate, y = dryMass)) +
  geom_point() +
  facet_wrap(vars(nlcdClass), ncol = 3) +
  xlab(expression(paste("Year")))+
  ylab(expression(paste("Dry Mass (g)"))) #check grams
print(Needles2) 

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: 7 because its easier to see the spread for the three different locations since they are not overlapped (like in 6). 